% !TeX root = ../main.tex

\chapter{Background and Related Work}
\label{ch:background}

This chapter provides the technical background on which the rest of the thesis builds. Section~\ref{sec:bg:wasm} reviews the WebAssembly execution model. Section~\ref{sec:bg:wasmedge} describes the WasmEdge runtime and its existing compile paths. Section~\ref{sec:bg:llvm-cost} characterizes the cost profile of LLVM-based compilation, which motivates the tiered architecture. Section~\ref{sec:bg:ir-lib} introduces the lightweight Sea-of-Nodes JIT library used as the tier-1 code generator. Section~\ref{sec:bg:tiered} summarizes the principles of tiered compilation. Section~\ref{sec:bg:related} surveys related work.

\section{The WebAssembly Execution Model}
\label{sec:bg:wasm}

WebAssembly is a low-level bytecode format designed for portable, sandboxed execution~\cite{haas2017wasm,wasm-spec}. A WebAssembly module is a binary container divided into typed sections: function signatures, function bodies, imports, exports, tables, linear memories, globals, and an initialization section for elements and data. The binary is validated in a single pass before any instruction executes. Validation enforces type safety on the operand stack, valid branch-target depths, and in-range indices for the function, local, global, table, and memory spaces.

WebAssembly defines a stack-based virtual machine. Each instruction either consumes typed operands from the operand stack and produces a typed result, or alters control flow. The instruction set covers integer and floating-point arithmetic, comparisons, memory load and store, function calls, and structured control. The MVP specification is augmented by feature proposals that this thesis targets: reference types, bulk-memory operations, and table operations. SIMD (\texttt{v128}) is not targeted in the current implementation; see Section~\ref{sec:conclusion:limitations}.

A module's defined functions, together with its imported functions, form an indexed function space. Direct calls reference these functions by index. Indirect calls reference a function pointer through a typed table, and the runtime checks the table entry's signature against the call site's expected signature at run time. Tables support host-controlled writes and runtime growth.

Linear memory is a byte-addressable buffer that grows in fixed-size pages. Load and store instructions specify a base address and a static offset. The runtime is required to bounds-check every access and to trap on a violation. Linear memory is not shared with the host process address space. This isolation is the foundation of WebAssembly's sandboxing guarantee.

Control flow is structured rather than free-form. Blocks, loops, and conditionals open scopes that can be exited only through labelled branch instructions, and every branch target is a syntactic structure visible from the branch instruction. The restriction simplifies validation and gives optimizing compilers a regular control-flow shape to operate on.

A module instance is a runtime materialization of a module against a specific store, with concrete function instances, table instances, memory instances, and global instances bound. Multiple module instances of the same module may coexist with independent state. Host functions---functions implemented by the embedder rather than by the wasm bytecode---are imported into the module instance through the same indexed function space as defined wasm functions. Call sites therefore do not distinguish between defined functions and host functions.

\section{The WasmEdge Runtime}
\label{sec:bg:wasmedge}

WasmEdge~\cite{wasmedge-docs} is an open-source WebAssembly runtime hosted by the Cloud Native Computing Foundation. The runtime targets cloud-native and edge deployments and is used in production by ByteDance, Microsoft, Sony, and others. Its codebase implements the full WebAssembly MVP specification, every standardized extension that the proposals working group has accepted, and several WASI standards.

WasmEdge provides three execution paths for a parsed and validated module. The interpreter walks the wasm bytecode directly, dispatching one instruction per loop iteration through a switch over the opcode. The interpreter is the simplest of the three paths and the slowest. It exists for correctness reference, for environments in which JIT-compiled code is unwelcome, and for kernels and instructions for which the compiled paths have not been implemented.

The two compiled paths share a common frontend. WasmEdge's LLVM frontend translates wasm bytecode to LLVM intermediate representation. It lowers wasm operations to typed LLVM operations, wasm structured control to LLVM basic blocks, and wasm linear memory accesses to LLVM loads and stores with explicit bounds-check guards. The frontend is reused by both compiled paths and is therefore the runtime's single source of wasm semantic lowering.

The ahead-of-time path runs the frontend at build time. The resulting LLVM module is compiled to a shared object that the runtime loads back at module instantiation. The shared object bypasses the frontend entirely on the loading thread, so cold start is dominated by file I/O and dynamic linking rather than by compilation. The path is intended for production deployments in which the wasm module is known in advance and can be compiled once.

The just-in-time path runs the same frontend at module instantiation. The resulting LLVM module is passed through LLVM's optimization pipeline and its just-in-time machine code emitter, and the produced function pointers are bound into the module instance. The path produces code of comparable quality to the ahead-of-time path. The path also pays the full LLVM compile cost on every cold start. That cost is the bottleneck this thesis addresses.

All three paths share the same module-instance representation and the same host-call dispatch boundary. A wasm-to-host call lands in a stub that marshals arguments from the wasm convention into the host's native convention and back. A host-to-wasm call goes through the runtime's executor to ensure that the trap state and stack guards are correctly maintained. The shared boundary is what allows this thesis to introduce a new compiled path---the tier-1 baseline JIT and selective tier-2 promotion---without changes to the executor's external interfaces.

\section{LLVM Compilation Cost}
\label{sec:bg:llvm-cost}

LLVM~\cite{lattner2004llvm} is a mature, production-grade optimizing compiler infrastructure. Its optimization pipeline at \texttt{O2} and above is comparable in coverage and aggressiveness to the optimizing tiers of major language compilers. The pipeline is also expensive. The cost has three sources.

The first source is the size of the pipeline itself. LLVM at \texttt{O2} runs dozens of analysis and transformation passes over the intermediate representation, including inlining, loop transformations, vectorization, and global value numbering. The passes operate on a single intermediate representation that is repeatedly rewritten as each transformation fires. Each pass adds constant overhead even on inputs it does not transform, because the analysis it depends on must still execute.

The second source is the size of the input. Compile time scales roughly linearly with the number of LLVM instructions, but specific passes scale superlinearly. Inlining and global value numbering are common offenders: their cost grows with both the number of call sites and the size of inlined bodies. A WebAssembly module with thousands of functions and tens of thousands of LLVM instructions per function exposes both kinds of growth.

The third source is memory footprint. The optimization pipeline holds the entire module in memory and rewrites it in place. Analyses build derived data structures that may rival the module itself in size. The footprint matters because edge deployments routinely run on hosts provisioned with one or two gigabytes of memory across the entire workload, of which compilation may consume a noticeable fraction.

The three sources combine to produce a compile-cost profile that is acceptable when paid once at build time on the AOT path, but is structurally unsuited to a cold-start path. The tiered architecture in Chapter~\ref{ch:design} reduces the up-front cost by routing the cold path through a much smaller code generator and reserving LLVM for the small subset of functions that benefit from its optimization budget.

\section{A Lightweight Sea-of-Nodes JIT Library}
\label{sec:bg:ir-lib}

This thesis adopts \emph{dstogov/ir}~\cite{dstogov-ir}, an open-source JIT compilation framework developed by Dmitry Stogov, the lead developer of the Zend Engine and the author of the Zend JIT shipped in PHP 8.0 and later. The library is a productized extraction of the same code generator that powers PHP's JIT path, designed from the outset to fit the compile-time budget of a JIT compiler rather than the compile-time budget of an ahead-of-time toolchain: a fixed linear-time pass pipeline, no LLVM-style pass-manager abstraction, and a linear-scan register allocator on SSA. These engineering choices yield compile-time profiles that scale linearly with function size and remain bounded under adversarial input, which is the property a baseline JIT requires on the cold path. The library's intermediate representation is Sea-of-Nodes in the sense of Click and Paleczny~\cite{click1995sea}: data and control dependencies are unified in a single graph of typed nodes, and most nodes are ``floating''---not pinned to any basic block---until a late scheduling pass places them. The same representation underlies HotSpot's C2 server compiler, V8's TurboFan, and Graal. Because the IR is SSA by construction, the library exposes a single-pass builder API that emits nodes one at a time and never needs a separate SSA-construction phase.

The optimization sequence is short and fixed. A folding engine runs during IR construction, performing constant folding, copy propagation, algebraic simplification, and common-subexpression elimination via value-numbered node deduplication. A sparse conditional-constant-propagation pass~\cite{dstogov-ir} folds branches and arithmetic on values that become constant under the SSA control flow. A global code motion pass after Click~\cite{click1995gcm} builds the CFG skeleton and pins each floating node to its best basic block, hoisting loop-invariant computations and sinking expressions out of hot blocks. A local scheduling pass then orders instructions within each block by topological sort. Target instruction selection performs Max-Munch tree pattern matching to combine adjacent nodes into single host instructions. A linear-scan register allocator on SSA assigns values to physical registers. The sequence is fixed: there is no pass manager, no opt-level abstraction, and no extension point for analyses that re-traverse the IR.

The fixed sequence is the source of the library's two attractive properties for a baseline JIT. Compile latency is bounded and predictable, because the cost of every pass scales linearly with the function size. The library's source code is small enough to vendor into the runtime and modify when a behavioural change is required. Both properties are essential to the tier-1 design described in Chapter~\ref{ch:design}: the runtime needs predictable compile cost on the cold path, and the thesis relies on the freedom to extend the library with backend changes that the project might not accept upstream.

The trade-off is code quality. The library does not perform inlining, does not vectorize, and does not run iterative dataflow analyses beyond the fixed sequence. Its output is competitive with a non-optimizing reference compiler and an order of magnitude below LLVM at \texttt{O2}. This gap is precisely the gap the tier-2 path closes for hot functions.

\section{Tiered Compilation Principles}
\label{sec:bg:tiered}

Tiered compilation is the design strategy of using two or more compilers of different cost-and-quality profiles to compile the same program, with profile information directing which compiler runs on which function~\cite{v8-wasm,webkit-wasm,firefox-tiering,wasmtime-baseline-rfc}. The strategy is established in production language runtimes for one consistent reason: the distribution of execution time across the functions of a typical program is highly skewed, and a single optimizing compiler spends most of its compile budget on functions that contribute little to run-time. The pattern generalizes beyond managed-language runtimes: Chen et al.~\cite{chen2024riscv} apply the same interpreter--baseline--optimizer layering to RISC-V instruction-set simulation, reporting that a lightweight baseline tier paired with a selective optimizing tier dominates either tier in isolation across a representative simulation workload.

A baseline tier produces native code for every defined function at low compile cost. The output is correct and reasonably fast, but does not benefit from heavy optimization. Common baseline-tier designs use a template-based code generator that emits per-opcode native sequences from a fixed pattern table, or a lightweight SSA-based code generator with a small fixed optimization sequence. The baseline tier is what eliminates the cold-start gap. The runtime can begin executing the program as soon as parsing and the baseline compile complete.

An optimizing tier produces code of significantly higher quality at significantly higher compile cost. The tier is engaged selectively. Only functions identified as worth the cost are recompiled, and the identification is driven by profile information collected from the baseline tier's executing code. The optimizing tier delivers steady-state performance comparable to a single-tier optimizing compiler, but only for code that actually runs hot.

Profile information takes the form of counters embedded in baseline-tier code. A per-function call counter increments on each entry to the function; when it crosses a threshold, the function is queued for optimizing-tier compilation. A per-loop back-edge counter increments on each loop iteration; when it crosses a threshold, the executing function is marked as containing a hot loop. The two counter classes target the two invocation shapes that matter: a frequently-called function and a long-running loop within a single invocation.

Promotion takes one of two forms. \emph{Function-entry promotion} swaps the dispatch entry for a function from its baseline-tier address to its optimizing-tier address, so that subsequent direct or indirect calls reach the optimized body. The currently-executing frame, if any, continues to run baseline-tier code, because the frame has already passed through the prologue. \emph{On-stack replacement}~\cite{fink2003osr,delia2018osr,holzle1992deopt} migrates an executing frame mid-function into an optimizing-tier continuation entry. The continuation entry is compiled with the loop header as the function entry point and the wasm-level state of the frame as input. On-stack replacement matters when a function is called once and spends nearly all its execution inside a loop; function-entry promotion delivers no speedup on such workloads.

The optimizing tier's compile cost must not block user code. Production tiered runtimes therefore run optimizing-tier compilation asynchronously on one or more worker threads. The user thread executes baseline-tier code, accumulates profile information, and notifies a worker on threshold crossings. The worker compiles and atomically publishes the result through a shared lookup table. Atomic publication ensures that concurrent callers either observe the baseline-tier pointer or the optimizing-tier pointer, never an intermediate state.

The architecture described in Chapter~\ref{ch:design} instantiates these principles for WasmEdge: a lightweight Sea-of-Nodes baseline tier, a selective recompilation tier driven by the existing LLVM frontend, profile counters embedded in baseline-tier bodies, both function-entry promotion and on-stack replacement, and an asynchronous worker thread with atomic publication into a shared dispatch table.

\section{Related Work}
\label{sec:bg:related}

The proposed architecture sits at the intersection of three established lines of work: tiered compilation in browser-engine WebAssembly pipelines, low-latency baseline compilation in server-side WebAssembly runtimes, and on-stack replacement as a runtime migration mechanism. This section surveys each, then states how the proposed design relates to them.

\subsection{Browser-engine WebAssembly tiering}
\label{sec:bg:related:browser}

The three major browser-engine WebAssembly pipelines all use multi-tier execution.

V8 pairs Liftoff with TurboFan~\cite{v8-wasm}. Liftoff is a single-pass baseline compiler that maps wasm operations directly to machine registers without constructing an intermediate representation, emitting code at a documented rate of tens of megabytes per second. TurboFan is V8's multi-pass optimizing compiler, invoked on hot functions identified by per-function execution counters. V8's documentation states that the pipeline does not perform on-stack replacement for WebAssembly: hot functions tier up at the next call, and any in-flight invocation runs to completion in Liftoff code. V8's more recent dynamic-tiering work~\cite{v8-dynamic-tiering} addresses the orthogonal cost of whole-module optimizing recompilation by deferring TurboFan to functions that prove hot at run time, rather than compiling the entire module eagerly.

SpiderMonkey pairs Baseline with Ion~\cite{firefox-tiering}. Baseline is the low-latency tier, reported as ten to fifteen times faster than Ion at compile time. Ion is the optimizing tier. Firefox's documented configuration historically performed eager whole-module Ion compilation after Baseline produced its initial output; subsequent work moved toward function-granularity Ion compilation triggered by warm-up counters.

JavaScriptCore now operates a three-tier WebAssembly pipeline~\cite{webkit-wasm}. The lowest tier is an interpreter (IPInt) that begins execution as soon as the module is parsed. BBQ (Build Bytecode Quickly) is the baseline JIT, documented as approximately four times faster at compile time than the optimizing tier. OMG (Optimized Machine-code Generator) is the optimizing JIT, invoked on hot functions. WebKit publishes call-site hot-patching as its function-promotion mechanism, and recent work has added interpreter-to-BBQ on-stack replacement; BBQ-to-OMG OSR is not documented in the public material.

The proposed design shares the startup-versus-throughput trade-off with all three pipelines and the selective hot-function promotion policy with V8's dynamic-tiering work. It differs in three respects. First, the baseline tier is built on a vendored Sea-of-Nodes code generator rather than on a template or single-pass emitter; the trade-off is discussed in Section~\ref{sec:bg:ir-lib}. Second, the optimizing tier reuses an existing wasm-to-LLVM frontend rather than a custom optimizing backend, which requires the mini-module synthesis and ABI bridge mechanism described in Chapter~\ref{ch:design}. Third, the architecture targets a server-side WebAssembly runtime; the baseline tier's compile-latency budget is tighter than Liftoff's or Baseline's, and the optimizing tier's compile-cost profile is heavier than TurboFan's, Ion's, or OMG's.

\subsection{Server-side WebAssembly runtimes}
\label{sec:bg:related:server}

Wasmtime~\cite{wasmtime-docs} is the closest server-side analogue to WasmEdge. Its default backend is Cranelift~\cite{cranelift}, an optimizing code generator designed to balance compile speed and code quality. The Wasmtime baseline-compilation RFC~\cite{wasmtime-baseline-rfc} introduces Winch (WebAssembly Intentionally Non-optimizing Compiler and Host) as a low-latency baseline that emits machine code directly from wasm bytecode in a single forward pass, using Cranelift's instruction emitters as a macro-assembler. Wasmtime's documentation states that the runtime does not currently support automatic tier-up from Winch to Cranelift: a module is compiled entirely by one backend or the other at the user's choice.

The proposed design differs from Wasmtime's current state in two ways. First, the architecture provides an automatic tier-up pathway: tier-2 promotion is driven by execution profile rather than a build-time strategy choice. Second, the design introduces on-stack replacement as part of the same tier-up pathway, addressing the workload shape that function-entry promotion alone cannot accelerate.

\subsection{On-stack replacement}
\label{sec:bg:related:osr}

On-stack replacement was introduced in the Self programming language's adaptive compilation pipeline as a mechanism for transferring an executing activation from one compiler's output into another's~\cite{holzle1992deopt}. The Self work used the mechanism in both directions: to enter optimized code from baseline code, and to recover from speculative optimizations that proved invalid at run time. Fink and Qian formalized OSR as the mechanism behind adaptive recompilation in the Jikes RVM~\cite{fink2003osr}, with stack maps generated by the baseline compiler describing the location of live variables at every potential transfer point and the optimizing compiler synthesizing prologue code that unpacks the stack map and reconstructs its own frame layout. D'Elia and Demetrescu provide a contemporary taxonomy of OSR mechanisms~\cite{delia2018osr}, distinguishing the transfer semantics (the correctness contract on values that must survive the transition) from the runtime engineering choices (stack-map representation, transfer-point granularity, continuation synthesis).

Modern production virtual machines treat OSR as a standard part of the tiering pathway. The Java HotSpot virtual machine has supported OSR for over two decades~\cite{paleczny2001hotspot}. A back-edge counter inside each compiled method triggers an OSR compile once the counter crosses a threshold. The OSR compile produces a second version of the method whose entry point is the loop header. The runtime then copies live local-variable state from the running frame into the new frame at the transfer point. The .NET CoreCLR runtime added the same mechanism in .NET 7, released in 2022~\cite{dotnet7osr}. The .NET release notes give the same motivation that drives this thesis: a method called once that spends most of its time inside a loop gets no benefit from function-entry promotion alone.

A separate family of runtimes uses trace compilation rather than OSR to enter optimized code from a running unoptimized frame. Tracing JIT compilers such as PyPy~\cite{bolz2009pypy} and LuaJIT start compiling when a back-edge counter crosses a threshold. The compiler records the actual execution path taken through the loop body. The recorded trace is then compiled as a code fragment whose entry point is the loop header. The interpreter can call into the trace directly on the next iteration, with no separate state-transfer step. Trace compilation is the main alternative to OSR-based tier-up in the literature. The approach is incompatible with this thesis's architecture: reusing the existing WasmEdge LLVM frontend requires a function-granularity compilation unit, not a trace-granularity one.

On-stack replacement also has a reverse direction. OSR-exit moves an executing optimized-tier frame back into a lower tier when an assumption the optimizer made turns out to be wrong at run time~\cite{holzle1992deopt}. HotSpot's ``uncommon trap'' mechanism and JavaScriptCore's FTL OSR-exit both implement this direction. The OSR-exit direction lets the optimizing tier make assumptions about the program that the baseline tier does not. This thesis implements only the OSR-entry direction. The absence of OSR-exit is discussed as a limitation in Chapter~\ref{ch:conclusion}.

The OSR mechanism described in Chapter~\ref{ch:design} follows the HotSpot and .NET-7 lineage outlined above and sits at the simpler end of the spectrum that D'Elia and Demetrescu describe. The transfer points are restricted to outermost-loop back-edges. The live state at each transfer point is exactly the set of wasm-level locals---a syntactically delimited and typed set rather than the register-and-spill location set that HotSpot-style stack maps must describe. The continuation is synthesized as a fresh wasm function whose entry point is the loop header. The simplification is made possible by WebAssembly's structured control flow and its explicit local-variable model. The price is a conservative filter that rejects loops whose surrounding control flow cannot be safely re-entered mid-execution; the rejection is discussed in Section~\ref{sec:design:osr:synthesis}.

\subsection{Position of this work}
\label{sec:bg:related:position}

Tiered compilation as a strategy is well established. The contribution of this thesis is not the strategy itself but a specific combination of design choices for a server-side WebAssembly runtime: a non-LLVM Sea-of-Nodes baseline tier inside a runtime whose existing optimizing backend is LLVM, a selective recompilation tier that reuses that LLVM frontend through per-batch mini-module synthesis rather than a parallel optimizing pipeline, an explicit ABI bridge layer that reconciles the two backends' calling conventions on every cross-tier dispatch path, and a loop-level OSR mechanism that complements function-entry promotion on the one-shot-caller workloads that dominate the architecture's target deployment. The combination is, to the best of the author's knowledge, not documented in either the browser-engine or the server-side WebAssembly literature in this exact form.
