% !TeX root = ../main.tex

\chapter{Conclusion}
\label{ch:conclusion}

\section{Summary}
\label{sec:conclusion:summary}

This thesis presents a tiered compilation architecture for WasmEdge. The architecture decouples cold-start latency from peak code quality by routing the cold path through a lightweight Sea-of-Nodes baseline JIT, then selectively recompiling hot functions with the existing WasmEdge LLVM frontend on a background worker thread. A profile-driven promotion mechanism replaces baseline-tier entries with optimizing-tier entries atomically; an on-stack replacement mechanism migrates in-flight tier-1 frames into optimizing-tier continuation entries mid-loop.

Three design contributions support the architecture. The first is a tier-1 baseline JIT that integrates a third-party lightweight Sea-of-Nodes JIT library into WasmEdge as a compile-every-function pathway. The tier embeds per-function and per-loop profile counters, and exposes a uniform two-argument calling convention that unifies wasm-defined functions, host imports, and indirect-call targets. The second is a tier-2 selective recompilation pipeline that reuses the existing LLVM frontend through per-batch mini-module synthesis. The pipeline introduces three ABI bridge thunks that reconcile the tier-1 and LLVM calling conventions, a compile-ordering invariant that preserves cross-tier call sites for post-processing rewriting, and a batch composition algorithm that walks up the static call graph from the threshold-tripping leaf and admits direct callees under a two-test gate. The third is an on-stack replacement mechanism that instruments outermost-loop back-edges with a three-state diamond and lifts wasm locals into the continuation entry's parameter list. The mechanism bails out structurally on loops whose surrounding control flow cannot be safely re-entered mid-execution.

The architecture is implemented in WasmEdge and evaluated on the sightglass-strong benchmark suite, a strengthened variant of upstream sightglass. Each kernel's tier-1 run-time lands in a 5- to 10-second band, giving tier-2 a measurable post-swap runway. The evaluation includes the suite's larger real-world workloads---the SpiderMonkey JavaScript engine on three input scripts, two TinyGo-compiled Go programs, SQLite, and two ONNX-runtime image classifiers---which expose the architecture to module sizes the synthetic kernels do not reach. Across the 39 analyzed kernels, the full configuration delivers a geometric-mean speedup of 1.11$\times$ over the tier-1 baseline and recovers approximately 86\% of whole-module LLVM JIT throughput, while reducing compile time by 11.1$\times$ on average. End-to-end wall-clock time---the metric that matters for cold-start workloads---improves by 1.73$\times$ on average over whole-module LLVM JIT. The largest wins concentrate on the real-world workloads, up to 21.9$\times$ on \texttt{tinygo-json} and 14.1$\times$ on the ONNX classifier. The step-by-step evaluation isolates the contribution of each mechanism: walk-up batch composition rescues the hot-leaf workloads (\texttt{shootout-ed25519} from 0.38$\times$ to 0.58$\times$, \texttt{hashset} from 0.62$\times$ to 0.84$\times$), and on-stack replacement closes the residual gap on one-shot-caller workloads (\texttt{shootout-ctype} from 0.37$\times$ to 0.88$\times$, \texttt{blind-sig} from 0.40$\times$ to 0.80$\times$, \texttt{shootout-random} from 0.63$\times$ to 1.00$\times$).

\section{Limitations}
\label{sec:conclusion:limitations}

The architecture has four structural limitations.

\paragraph{No SIMD lowering.} Tier-1 does not lower WebAssembly SIMD instructions: \texttt{v128} loads and stores, lane operations, shuffles, and the per-lane arithmetic and comparison families. Modules whose compiled wasm contains these instructions cannot be loaded through the tiered pipeline at all---the translation pass rejects them at module instantiation, in the same way that \texttt{splay}'s garbage-collection instructions are rejected. Cryptographic, vectorized hashing, and image-processing workloads that ship as SIMD-compiled wasm (\texttt{libsodium}, \texttt{intgemm-simd}, \texttt{meshoptimizer}, \texttt{blake3-simd}, \texttt{hex-simd}) are therefore outside the current evaluation scope; where a scalar variant exists (\texttt{blake3-scalar}), it is used instead. A related scalar-only filter sits at the tier-2 promotion gate: regular batches and OSR alike only admit functions whose signature is entirely scalar (\texttt{i32}, \texttt{i64}, \texttt{f32}, \texttt{f64}, with at most one return). Functions whose parameters or return types include \texttt{v128}, reference types, or multi-return signatures stay on the tier-1 path.

\paragraph{No de-tiering.} Once a function or loop is promoted to tier-2, the architecture provides no path back to tier-1. The decision is one-way. For long-running modules in which the workload shape changes over time, this design choice leaves no recovery option if a tier-2 specialization becomes a poor fit for a later phase of execution. The choice reflects a simplicity-versus-completeness trade-off appropriate to the cold-start use case the thesis targets, where the workload is typically short-lived.

\paragraph{x86-64 host support only.} The tier-1 backend, the OSR continuation synthesis, and the ABI bridge thunks are all implemented and tested on x86-64. The underlying backend supports aarch64, and the thesis's modifications are largely architecture-neutral, but no aarch64 port has been validated. Edge deployments on ARM-based hosts are therefore not yet covered.

\paragraph{Worker-thread CPU contention on small hosts.} The evaluation runs on a 24-core workstation in which the tier-2 worker thread executes on a physical core that the user thread never contends for. The compile-time costs reported in Chapter~\ref{ch:evaluation} are therefore off the critical path of the user thread. Edge nodes are frequently provisioned with one or two vCPUs, at which point the worker thread shares time with the user thread under the host scheduler. A long LLVM compile on a single-vCPU host will preempt the user thread for the duration of the compile, eroding the wall-clock benefit that the architecture targets. The mechanism is not broken in this regime---tier-1 execution still proceeds, and the swap still fires correctly when the worker eventually completes---but the cold-start latency win is bounded by the host's ability to schedule the worker alongside, rather than in place of, useful execution. Mitigations are deployment-side rather than architecture-side: setting the worker thread to a lower scheduling priority, constraining the worker's CPU share through cgroups or affinity masks, or yielding cooperatively at coarse-grained LLVM-pass boundaries. None of these are implemented in the current prototype, and the appropriate policy depends on host characteristics that the runtime does not currently introspect.

\section{Future Work}
\label{sec:conclusion:future}

Four directions extend the architecture.

\paragraph{SIMD lowering and reference-type promotion scope.} Adding tier-1 lowering for the WebAssembly SIMD (\texttt{v128}) instruction set is the primary missing piece. The lowering itself is mechanical---each SIMD opcode maps to one or a small group of IR backend SSE/AVX nodes---but the local-serialization buffer used by the OSR transition must also be widened to a type-native \texttt{v128} slot, and the interaction with the IR backend's common-subexpression elimination (which dictates the type-native-width design in the scalar case) must be re-validated for the wider type. Once tier-1 lowers these instructions, the scalar-only filter at the tier-2 promotion gate can be relaxed to admit \texttt{v128} and reference-type signatures, and the OSR transition buffer can be extended to marshal the additional types across the swap.

\paragraph{aarch64 port.} Validating the tier-1 backend, the ABI bridge thunks, and the OSR continuation synthesis on aarch64 brings ARM-based edge hosts into scope. The bridge thunks must be re-derived for the AAPCS64 calling convention, and the OSR continuation entry's parameter layout must be re-validated against the aarch64 backend's register allocator.

\paragraph{Refcounted JIT cache for de-tiering.} Adding a refcounted cache for tier-1 native code, indexed by function index and held alive while any frame may still observe the code, opens a path to de-tiering. A function or loop incorrectly promoted under one workload shape could be reverted to tier-1 once the workload shape changes, and re-promoted later with updated profile information.

\paragraph{Profile-guided batch policy.} The current batch composition uses a fixed walk-up depth and a fixed BFS bound. Profile information beyond the per-function and per-loop counters---call-site frequencies, branch biases, value profiles---could drive a more selective batch policy that includes the surrounding context only when the profile indicates it will be inlined or specialized to advantage. The expected gain is on workloads where the current batch is too small (kernels still below 0.80$\times$ LLVM JIT after Step 3) and where the current batch is too large (kernels where the worker spends compile cycles on cold paths that the optimizer cannot prune).
