% !TeX root = ../main.tex

\chapter{Evaluation}
\label{ch:evaluation}

This chapter evaluates the tiered compilation architecture. Section~\ref{sec:eval:method} introduces the benchmark suite, describes the strengthening procedure, defines the measurement protocol, and lists the hardware and parameters. Section~\ref{sec:eval:workloads} groups the kernels by execution shape; each tier-2 mechanism then maps to the group it targets. Section~\ref{sec:eval:llvm} establishes the whole-module LLVM JIT baseline as the upper bound on steady-state throughput. Sections~\ref{sec:eval:step1}--\ref{sec:eval:step3} measure the tiered architecture in three configurations, one per design step. Section~\ref{sec:eval:tt} reports compile time and end-to-end wall-clock time.

\section{Methodology}
\label{sec:eval:method}

\subsection{The sightglass benchmark suite}
\label{sec:eval:method:sg}

Sightglass~\cite{sightglass} is the WebAssembly benchmark suite maintained by the Bytecode Alliance. The suite ships a collection of kernels drawn from cryptographic primitives, parsing and serialization workloads, image and string processing, hashing, and small synthetic compute loops. Each kernel is compiled to WebAssembly through the standard toolchain and packaged with an input file and a golden output. The kernels are diverse enough to expose backend code-quality differences, and small enough that each kernel completes in seconds rather than minutes. This thesis uses Sightglass as an internal harness for comparing successive configurations of one runtime. It is not used as a cross-runtime benchmark. No claim of universal representativeness is made on its behalf.

The kernels are designed for cheap regression-testing. Most upstream kernels finish in well under a second on a modern host. The brevity is appropriate for correctness regression. The brevity is a measurement hazard for tier-2 evaluation. Tier-2 works by executing under the tier-1 baseline while a worker thread compiles the LLVM-grade replacement, then switching to the optimized code once it is ready. If the entire kernel completes before the switch, the measurement reflects only tier-1 work and reveals nothing about tier-2 code quality. Even when the switch fires, a residual tier-1 prefix of comparable size to the post-switch tail confounds the steady-state signal.

\subsection{The sightglass-strong suite}
\label{sec:eval:method:strong}

To give tier-2 a measurable runway, this thesis strengthens every short kernel so that its tier-1-only run-time lands in a 5- to 10-second band, with a target near 8 seconds. The strengthened suite is referred to as \emph{sightglass-strong}. A few seconds of post-switch execution is enough to amortize the worker's compile cost and to distinguish real steady-state speedup from measurement noise.

Three strengthening patterns cover most kernels. The first pattern increases an internal iteration constant defined at the top of the kernel source, used for shootout-style kernels whose inner loop is bounded by a single macro. The second pattern bumps the literal in the kernel's outer benchmark loop, used for kernels whose driver already loops over the work region. The third pattern wraps the benchmark region in a fresh outer loop, used for Rust and C kernels that originally measured a single execution. A small number of kernels require additional work: out-of-bounds fixes uncovered by the larger iteration count, optimization barriers to prevent the compiler from hoisting the kernel out of the surrounding loop, and replacement of nondeterministic output (such as wall-clock prints) so that the golden output remains stable. The per-kernel strengthening table is reproduced in Appendix~A.

The evaluation also includes eight large real-world workloads that ship with upstream sightglass but sit outside its default suite: \texttt{spidermonkey-json}, \texttt{spidermonkey-markdown}, \texttt{spidermonkey-regex} (the SpiderMonkey JavaScript engine compiled to WebAssembly, running three different input scripts), \texttt{tinygo-json}, \texttt{tinygo-regex} (TinyGo-compiled Go programs), \texttt{image-classification} and \texttt{tract-onnx-image-classification} (ONNX-runtime image classifiers), and \texttt{sqlite3} (the SQLite database engine). These workloads expose the tiered architecture to module sizes and compile-cost profiles that the synthetic kernels do not reach. The full strengthened suite contains 41 measured kernels.

Goldens for sightglass-strong were regenerated by running the strengthened kernels under an independent runtime---wasmtime 43.0.1---rather than under WasmEdge itself. The independent oracle means that any regression in WasmEdge's parser, baseline JIT, or tier-2 pipeline surfaces as a golden mismatch. Without an independent oracle, the runtime output and the reference output would corrupt together silently.

\subsection{Measurement protocol}
\label{sec:eval:method:protocol}

Each measurement reports a three-run median to reduce single-run variance. All measurements are taken on the same host with the runtime built in release-optimized configuration and the tier-1 backend at its standard optimization level. Three metrics are reported. \emph{WorkTime} (\texttt{WT}) is the harness-measured execution time of the benchmark region. \emph{Compile time} (\texttt{Comp}) is the time spent compiling wasm functions to native code. The tier-1 path pays this cost before the first wasm instruction runs; the tier-2 path schedules it on the worker thread. \emph{Time-to-value} (\texttt{TtV}) is the end-to-end wall-clock time from process start to benchmark completion (\texttt{Comp + WT} to a close approximation).

These three metrics map to the vocabulary used in the serverless and edge literature~\cite{gackstatter2022serverless}. \emph{Cold-start latency} is the time from request arrival at the runtime to the start of execution of the first machine-code instruction of the requested wasm function, on the assumption that no ready-to-run compiled instance of the module is already resident on the host. Under this definition, the full cold-start latency includes file I/O, parse, validate, compile, and instantiate. Compile dominates this sum on the kernels in the evaluation suite. The non-compile portion---load, parse, validate, and instantiate---is mode-independent: it runs the same code path under whole-module LLVM JIT and under the tiered IR-JIT path, contributes equally to both, and cancels out of every cross-mode ratio. This thesis therefore uses \texttt{Comp} as the operational measure of cold-start latency throughout the evaluation. The end-to-end wall-clock metric \texttt{TtV} is the composite of cold-start latency and steady-state execution; this thesis refers to it as ``end-to-end wall-clock time'' or \texttt{TtV}, never as ``latency'' alone, to avoid confusion with the cold-start sense.
All ratios in this chapter follow the convention that values greater than one indicate a tier-2 win. The two ratios cited most often are \texttt{t1/t2} (tier-1 work-time divided by tier-2 work-time) and \texttt{LLVM/t2} (whole-module LLVM JIT work-time divided by tier-2 work-time). The configuration of interest is the second. The ratio \texttt{LLVM/t2} measures how close the tiered architecture comes to delivering whole-module LLVM JIT's steady-state throughput. The same convention applies to \texttt{Comp} and \texttt{TtV}: a ratio above one means tier-2 is faster. Per-kernel raw measurements for every configuration are reproduced in Appendix~B.

\subsection{Experiment setup}
\label{sec:eval:method:setup}

All measurements were taken on a single workstation-class host with no other interactive workload running. The host has a 13th-generation Intel Core i9-13900K processor (24 physical cores spanning performance and efficiency partitions, 32 hardware threads, 5.8\,GHz turbo) and 128\,GiB of DDR5 memory, running Ubuntu 22.04 LTS on Linux kernel 6.8. Each sightglass-strong run executes single-threaded. The worker thread that runs tier-2 compilation is the only other thread of interest. The worker is co-located with the user thread on the same host, but never contends for the same core during the benchmark region.

WasmEdge is built in release-optimized configuration. The tier-1 backend runs at its standard optimization level, which is the level the rest of the IR JIT pipeline uses. The whole-module LLVM JIT baseline runs through the unmodified WasmEdge LLVM frontend at its default optimization level, which is the level used for the production AOT path.

Three configurations are measured. Each configuration shares the same tier-1 baseline; the configurations differ only in which tier-2 mechanisms are enabled. Table~\ref{tbl:eval:params} summarizes the parameters.

\begin{table}[h]
\centering
\small
\begin{tabular}{l p{3.5cm} p{3.5cm} p{3.0cm}}
\toprule
Configuration & Tier-2 promotion & Function-entry threshold & OSR threshold \\
\midrule
Tier-1 only           & disabled                       & ---       & --- \\
Step 1 (hot callees)  & enabled, leaf + direct callees & 10 calls  & legacy per-loop counter \\
Step 2 (walk-up + BFS) & enabled, walk-up + BFS        & 10 calls  & OSR disabled \\
Step 3 (+ OSR)        & enabled, walk-up + BFS         & 10 calls  & 5000 iterations \\
LLVM JIT baseline     & not applicable                 & ---       & --- \\
\bottomrule
\end{tabular}
\caption{Configurations used in this evaluation. The function-entry threshold is the number of calls a tier-1 function counts before triggering tier-2 promotion. The OSR threshold is the number of back-edge iterations a tier-1 loop counts before triggering an OSR continuation compile.}
\label{tbl:eval:params}
\end{table}

The two thresholds are kept fixed across kernels. Step 1 corresponds to the earliest tier-2 design measured on the \texttt{promote-hot-callees} branch (commit \texttt{f07da860}); Steps 2 and 3 correspond to the current head on the \texttt{osr} branch (commit \texttt{bdd80e91}). The tier-1 baseline numbers used in every step come from the current head, so the step-to-step deltas reflect changes to the tier-2 pipeline rather than incidental changes to tier-1.

The two threshold values are calibrated to the call-count and iteration-count distributions of the strengthened suite. The strengthening procedure tunes each kernel to a 5--10 second tier-1 work-region, which forces every kernel's hot functions to execute many thousands to many millions of times during measurement. A function-entry threshold of 10 therefore saturates within milliseconds of kernel start on every kernel, while remaining well above the call counts of one-time initialization, validation, and cleanup helpers that should not be queued. The OSR threshold sits an order of magnitude below the shortest one-shot-caller loop in the suite. The relevant iteration counts range from 60{,}000 (\texttt{shootout-ctype}, the shortest) to two billion (\texttt{shootout-random}), passing through 1.2 million (\texttt{shootout-base64}), 35 million (\texttt{shootout-matrix}), and 50 million (\texttt{shootout-ratelimit}). A threshold of 5000 fires on every one of them. At most 8\% of the loop runs in tier-1 on \texttt{shootout-ctype}, and less than one part in ten thousand on \texttt{shootout-random}. The threshold also sits above the iteration counts of short bookkeeping loops, so the worker queue does not fill with continuations the user never re-enters. Both choices are robust over roughly a decade in either direction; the per-kernel statistics are several orders of magnitude away from the threshold on both sides.

\subsection{Excluded kernels}
\label{sec:eval:method:excluded}

Three kernels in the strong suite are excluded from the geometric means reported in this chapter. The kernel \texttt{noop} is excluded because its sub-microsecond run-time makes any ratio with seconds-scale kernels meaningless. The kernel \texttt{shootout-ackermann} is bimodal on the test host: depending on dynamic factors that have not been fully isolated, it lands in one of two well-separated time bands per run. The bimodality dominates per-kernel ratios involving this kernel. The kernel \texttt{splay} from upstream sightglass is not measured at all, because it uses WebAssembly garbage-collection instructions that tier-1 does not yet support; the kernel fails at tier-1 translation. The remaining 39 kernels constitute the analysis set.

A second class of upstream sightglass kernels is excluded one level earlier, before the strong suite is even assembled: kernels whose compiled wasm uses SIMD (\texttt{v128}) instructions, which tier-1 does not lower. The excluded directories are \texttt{blake3-simd}, \texttt{hex-simd}, \texttt{intgemm-simd}, \texttt{meshoptimizer}, and \texttt{libsodium} (the AEGIS, AES-GCM, and salsa20 families). For \texttt{blake3}, a scalar variant (\texttt{blake3-scalar}) is included in its place; the other four upstream directories have no scalar variant, so excluding the directory excludes the kernel from this evaluation entirely. The limitation is discussed in Section~\ref{sec:conclusion:limitations}.

\section{Workload Characterization}
\label{sec:eval:workloads}

The 39 analysis kernels divide into five execution-shape categories. The categorization clarifies which tier-2 mechanism each kernel responds to. The categories are not strictly disjoint: two kernels (\texttt{shootout-ctype} and \texttt{blind-sig}) sit in two categories simultaneously, and the design's full effect on those kernels requires both mechanisms to engage.

\paragraph{Frequent-call workloads.} The kernel invokes a hot function many times, with each call short relative to the total run-time. The function's entry counter saturates quickly, the entry-table swap captures every subsequent call, and the tier-1 prefix preceding the swap is negligible relative to the steady-state tail. Members: \texttt{blake3-scalar}, \texttt{bz2}, \texttt{gcc-loops}, \texttt{hashset}, \texttt{pulldown-cmark}, \texttt{quicksort}, \texttt{regex}, \texttt{shootout-gimli}, \texttt{shootout-heapsort}, \texttt{shootout-keccak}, \texttt{shootout-memmove}, \texttt{shootout-seqhash}, \texttt{shootout-sieve}, \texttt{shootout-switch}, \texttt{shootout-xblabla20}, \texttt{shootout-xchacha20}, \texttt{tinygo-regex}.

\paragraph{Hot-leaf workloads.} The threshold-tripping function sits several layers below the actual hot work. A small leaf helper crosses its counter first, but the surrounding caller---which contains the bulk of the run-time---is what tier-2 needs to compile to deliver speedup. Promoting only the leaf yields a tier-2 body that calls back into tier-1 for every helper outside the batch, blocking inlining and pinning the call sites at bridge-thunk cost. Members: \texttt{blind-sig}, \texttt{rust-compression}, \texttt{rust-json}, \texttt{rust-protobuf}, \texttt{shootout-ctype}, \texttt{shootout-ed25519}.

\paragraph{One-shot caller workloads.} The kernel invokes its outer function exactly once and spends nearly all of its run-time inside one long-running loop. Function-entry promotion delivers no speedup on these kernels, because the function never returns to its prologue and never re-enters through the swapped slot. Members: \texttt{blind-sig}, \texttt{shootout-base64}, \texttt{shootout-ctype}, \texttt{shootout-matrix}, \texttt{shootout-minicsv}, \texttt{shootout-random}, \texttt{shootout-ratelimit}.

\paragraph{Resistant workloads.} The kernel has either too little total run-time for the worker to amortize its compile cost, or structural properties that block optimizer wins. Members: \texttt{richards} (short total run-time), \texttt{rust-html-rewriter} (short total run-time), \texttt{shootout-fib2} (pure recursive call), \texttt{shootout-nestedloop} (innermost loop with no inline opportunity).

\paragraph{Compile-dominated workloads.} The kernel's LLVM compile cost dwarfs its WebAssembly work-time. Steady-state throughput is largely irrelevant on these workloads; end-to-end wall-clock time is dominated by how long the runtime spends in compilation before the program produces useful output. Members: \texttt{image-classification}, \texttt{spidermonkey-json}, \texttt{spidermonkey-markdown}, \texttt{spidermonkey-regex}, \texttt{sqlite3}, \texttt{tinygo-json}, \texttt{tract-onnx-image-classification}. These are the workloads for which the cold-start architecture exists.

\section{LLVM JIT Baseline}
\label{sec:eval:llvm}

Whole-module LLVM JIT establishes the upper bound on steady-state throughput. Its work-time numbers are the divisor for every \texttt{LLVM/t2} ratio reported in this chapter. The baseline also defines the cost structure that the tiered architecture is designed to avoid.

LLVM JIT compiles every defined function before any wasm instruction executes. Its compile time therefore scales with module size and the complexity of individual function bodies. Three groups appear in the LLVM JIT compile-time distribution. Small modules with few or simple functions compile in tens to hundreds of milliseconds (\texttt{shootout-gimli} 17 ms, \texttt{shootout-heapsort} 82 ms, \texttt{shootout-keccak} 517 ms). Mid-sized modules compile in one to five seconds (\texttt{rust-html-rewriter} 4.7\,s, \texttt{regex} 5.0\,s, \texttt{rust-compression} 5.6\,s). The eight large workloads added to the suite push the cost into the tens of seconds: \texttt{tinygo-regex} 7.8\,s, \texttt{tinygo-json} 10.3\,s, \texttt{sqlite3} 16.4\,s, and the three SpiderMonkey kernels at approximately 90\,s each. The ONNX-runtime classifier (\texttt{tract-onnx-image-classification}) compiles in 136\,s.

End-to-end wall-clock (\texttt{TtV}) is the metric that matters for cold-start workloads. LLVM JIT's TtV includes its compile time in full. For the compile-dominated kernels, LLVM JIT's compile time is one to four orders of magnitude larger than its work time: \texttt{spidermonkey-json} compiles for 90.7\,s and runs for 0.07\,s; \texttt{tinygo-json} compiles for 10.3\,s and runs for 0.05\,s. The user waits a minute and a half for SpiderMonkey to start; the actual JavaScript execution finishes in under a tenth of a second. The tiered architecture's design goal is to reduce TtV on these workloads without losing more steady-state throughput than is necessary on the long-running kernels.

\section{Step 1: Promote Hot Callees}
\label{sec:eval:step1}

The earliest tier-2 design promotes only the function whose counter saturated, together with its direct callees. The batch's root is the threshold-tripper itself; the breadth-first traversal extends one hop to its direct callees. There is no walk up the call graph.

Across the 39 analyzed kernels, the geometric mean of \texttt{t1/t2} is \textbf{0.97$\times$} and the geometric mean of \texttt{LLVM/t2} is \textbf{0.76$\times$}. Tier-2 in this configuration is on average slightly slower than tier-1, and recovers approximately 76\% of LLVM JIT throughput.

The aggregate hides two distinct failure modes.

\paragraph{Hot-leaf failure.} For kernels in the hot-leaf category, the threshold-tripper is a small helper. Compiling only the leaf yields a tier-2 body whose call sites all dispatch back into tier-1 through the ABI bridge thunks introduced in Section~\ref{sec:design:context:thunks}. The optimizer cannot inline across the thunk boundary, so the calls remain expensive. The kernels in this category show the largest gaps to LLVM JIT: \texttt{shootout-ctype} at 0.31$\times$, \texttt{blind-sig} at 0.37$\times$, \texttt{shootout-ed25519} at 0.38$\times$. Each is below 40\% of LLVM JIT throughput.

\paragraph{One-shot-caller failure.} For kernels in the one-shot-caller category, the entry-table swap fires after the outer function has already started running. The in-flight invocation runs the entire loop in tier-1 and returns. The swap is observable only on subsequent calls, and those calls never happen. \texttt{shootout-ctype}, \texttt{blind-sig}, \texttt{shootout-random}, and \texttt{shootout-base64} all sit well below the geometric mean for this reason. Several kernels also regress relative to tier-1. The worker's compile cost is paid without a sufficient post-swap tail to amortize it: \texttt{hashset} runs at 0.73$\times$ \texttt{t1/t2}, \texttt{shootout-ed25519} at 0.66$\times$, \texttt{shootout-ctype} at 0.54$\times$.

The two improvements described in the following sections each address one of these failure modes.

\section{Step 2: Walk-Up and BFS Batch Composition}
\label{sec:eval:step2}

The improved batch composition walks up the static call graph from the threshold-tripper to find the hottest direct caller, then performs a bounded breadth-first traversal of that caller's direct callees. Callee inclusion is gated by the two-test admission rule described in Section~\ref{sec:design:tier2:batch}. The walk-up rolls the surrounding caller into the batch, so the optimizer has full bodies to inline into.

Across the 39 kernels, the geometric mean of \texttt{t1/t2} rises to \textbf{1.03$\times$} and the geometric mean of \texttt{LLVM/t2} rises to \textbf{0.80$\times$}. Tier-2 now beats tier-1 by approximately 3\% on average and recovers approximately 80\% of LLVM JIT throughput.

The largest beneficiaries are kernels whose Step 1 batch was structurally too small for the optimizer to inline within (Table~\ref{tbl:eval:walkup}). \texttt{hashset} rises from 0.62$\times$ to 0.84$\times$, recovering from a Step 1 \texttt{t1/t2} regression to a clear tier-2 win. \texttt{shootout-ed25519} rises from 0.38$\times$ to 0.58$\times$ as the walk-up captures the cryptographic caller that wraps the modular-arithmetic leaf. The Rust kernels (\texttt{rust-json}, \texttt{rust-compression}, \texttt{rust-protobuf}) recover similarly: their static call graphs are deeper than the shootout kernels, so a single-hop BFS at Step 1 misses most of the hot work.

\begin{table}[h]
\centering
\small
\begin{tabular}{l r r r}
\toprule
Kernel & Step 1 \texttt{LLVM/t2} & Step 2 \texttt{LLVM/t2} & Swing \\
\midrule
\texttt{hashset}           & 0.62$\times$ & 0.84$\times$ & $+0.22$ \\
\texttt{shootout-ed25519}  & 0.38$\times$ & 0.58$\times$ & $+0.20$ \\
\texttt{rust-json}         & 0.62$\times$ & 0.75$\times$ & $+0.13$ \\
\texttt{rust-compression}  & 0.75$\times$ & 0.87$\times$ & $+0.12$ \\
\texttt{regex}             & 0.85$\times$ & 0.95$\times$ & $+0.10$ \\
\texttt{pulldown-cmark}    & 0.75$\times$ & 0.84$\times$ & $+0.09$ \\
\texttt{gcc-loops}         & 0.74$\times$ & 0.83$\times$ & $+0.09$ \\
\bottomrule
\end{tabular}
\caption{Kernels with the largest \texttt{LLVM/t2} gain when the batch composition switches from leaf-only to walk-up plus BFS (Step 1 $\to$ Step 2).}
\label{tbl:eval:walkup}
\end{table}

The frequent-call kernels that were already in good shape at Step 1 show small changes. \texttt{shootout-keccak} moves from 0.97$\times$ to 0.98$\times$, \texttt{shootout-gimli} from 0.99$\times$ to 0.99$\times$, \texttt{blake3-scalar} from 0.96$\times$ to 0.99$\times$. Function-entry promotion was already capturing the hot path on these kernels, and the larger batch the optimizer sees does not add additional inlining opportunities.

Two kernel groups remain well below LLVM JIT throughput at Step 2. The first is the one-shot-caller group: \texttt{shootout-ctype} reaches only 0.37$\times$, \texttt{blind-sig} only 0.40$\times$, \texttt{shootout-random} only 0.63$\times$. The function-entry promotion that walk-up plus BFS produces is correct and fires. It simply has no effect on the in-flight invocation. For these kernels, the in-flight invocation is the only invocation. Closing this group is the role of on-stack replacement. The second is the compile-dominated group: \texttt{spidermonkey-regex} reaches 0.55$\times$, \texttt{tract-onnx-image-classification} 0.63$\times$, \texttt{sqlite3} 0.68$\times$. The steady-state gap on these workloads is structural: the hot paths use vector and floating-point patterns that the tier-1 backend lowers without LLVM-grade scheduling. Their end-to-end wall-clock time is the relevant metric instead, and it tells a different story (Section~\ref{sec:eval:tt}).

\section{Step 3: On-Stack Replacement}
\label{sec:eval:step3}

The full configuration adds on-stack replacement (OSR) on top of walk-up and BFS. Tier-1 code now instruments each outermost-loop back-edge with the three-state diamond described in Section~\ref{sec:design:osr:diamond}. When the back-edge counter for a loop saturates, the worker thread compiles a continuation entry whose entry point is the loop header. The currently-executing tier-1 frame migrates into the continuation entry on the next iteration.

Across the 39 kernels, the geometric mean of \texttt{t1/t2} rises to \textbf{1.11$\times$} and the geometric mean of \texttt{LLVM/t2} rises to \textbf{0.86$\times$}. Tier-2 with OSR beats tier-1 by approximately 11\% on average and recovers approximately 86\% of LLVM JIT throughput.

The one-shot-caller kernels show the largest swings (Table~\ref{tbl:eval:osr}). The seven kernels with the biggest \texttt{LLVM/t2} jump from Step 2 to Step 3 are all from this category. The two kernels at the top of the table---\texttt{shootout-ctype} and \texttt{blind-sig}---swing more than half of the gap between Step 2 and LLVM JIT. Both kernels sit in both the hot-leaf and one-shot-caller categories. Walk-up at Step 2 produced a large optimizing-tier batch for each. The in-flight invocation never reached the swapped entry, so the batch was wasted. OSR delivers the in-flight migration that closes the remaining gap.

\begin{table}[h]
\centering
\small
\begin{tabular}{l r r r}
\toprule
Kernel & Step 2 \texttt{LLVM/t2} & Step 3 \texttt{LLVM/t2} & Swing \\
\midrule
\texttt{shootout-ctype}    & 0.37$\times$ & 0.88$\times$ & $+0.51$ \\
\texttt{blind-sig}         & 0.40$\times$ & 0.80$\times$ & $+0.40$ \\
\texttt{shootout-random}   & 0.63$\times$ & 1.00$\times$ & $+0.37$ \\
\texttt{shootout-base64}   & 0.78$\times$ & 0.99$\times$ & $+0.21$ \\
\texttt{shootout-matrix}   & 0.75$\times$ & 0.95$\times$ & $+0.20$ \\
\texttt{shootout-ratelimit}& 0.79$\times$ & 0.96$\times$ & $+0.17$ \\
\texttt{shootout-minicsv}  & 0.79$\times$ & 0.95$\times$ & $+0.16$ \\
\bottomrule
\end{tabular}
\caption{Kernels with the largest \texttt{LLVM/t2} gain when OSR is enabled (Step 2 $\to$ Step 3). All seven are from the one-shot-caller category.}
\label{tbl:eval:osr}
\end{table}

The mechanism behind each swing is identical. The outer function is called once. Almost all run-time is spent in one long-running loop. Without OSR, the function-entry slot swap fires after the loop has already started and is therefore observable only on a subsequent call that never happens. With OSR, the loop's back-edge counter saturates while the loop is still running, the worker compiles a continuation entry, and the tier-1 frame jumps into the continuation entry on the next iteration. The remainder of the loop runs in optimizing-tier code.

A small number of kernels regress when OSR is enabled. \texttt{rust-json} drops from 0.75$\times$ to 0.63$\times$, \texttt{spidermonkey-json} from 0.80$\times$ to 0.70$\times$, \texttt{pulldown-cmark} from 0.84$\times$ to 0.78$\times$, \texttt{rust-html-rewriter} from 0.67$\times$ to 0.62$\times$. These kernels share two properties: the inner loops are short relative to the rest of the program, and the hot functions are called frequently enough that function-entry promotion at Step 2 was already capturing the hot path. The added back-edge instrumentation costs three operations per iteration on the post-saturation steady state, which is small in absolute terms but visible when the loop body itself is short. The OSR continuation compile also competes with regular tier-2 batches for worker time. The OSR-priority drain (Section~\ref{sec:design:tier2:worker}) ensures continuation entries land before the loop exits on the kernels for which OSR is the only mechanism that helps; the cost is a small delay for regular-tier-2 compiles on kernels for which OSR is unnecessary.

\section{Compile Time and End-to-End Wall-Clock Time}
\label{sec:eval:tt}

The work-time analysis above is the steady-state metric. The compile metric is \texttt{Comp}; the end-to-end metric is \texttt{TtV}.

Across the 39 kernels, the geometric mean of \texttt{Comp T2/T1} in the full configuration is \textbf{3.05$\times$}. The worker thread takes about three times as long as tier-1 alone before it produces a tier-2 body for the hot batch. The absolute gap is tens to hundreds of milliseconds on a typical kernel. The bulk of optimization runs off the critical path on the worker thread, so this is the cost the kernel pays for tier-2 compilation, not the cost the user thread observes. The ratio of LLVM JIT compile time to tier-2 compile time, \texttt{Comp LLVM/T2}, has a geometric mean of \textbf{11.1$\times$}. Tier-2 finishes its compilation approximately eleven times faster than whole-module LLVM JIT on average. On the largest kernels the ratio is dominated by the absolute size of LLVM's compile work; on the smallest kernels the ratio is dominated by the fixed overhead of LLVM's pass pipeline.

End-to-end \texttt{TtV} integrates the trade-off. Across the 39 kernels, the geometric mean of \texttt{TtV LLVM/T2} is \textbf{1.73$\times$}. The tiered architecture finishes the end-to-end workload approximately 73\% faster than whole-module LLVM JIT on average, even though its steady-state throughput is 14\% lower (0.86$\times$ \texttt{LLVM/t2}). The gain comes from the saved compile time.

The benefit is largest on the compile-dominated workloads. \texttt{tinygo-json} finishes in 473 ms under tier-2 versus 10{,}361 ms under LLVM JIT, a 21.9$\times$ wall-clock win. \texttt{tract-onnx-image-classification} finishes in 9.7 s under tier-2 versus 136 s under LLVM JIT, a 14.1$\times$ win. The three SpiderMonkey kernels each finish in approximately 9 s under tier-2 versus approximately 90 s under LLVM JIT, a 10$\times$ win. \texttt{sqlite3} finishes 8.5$\times$ faster, \texttt{image-classification} 8.8$\times$ faster. These are exactly the workloads the cold-start architecture targets: real applications and large modules where the user waits seconds to minutes for LLVM JIT to start, while the actual benchmark region takes milliseconds to seconds.

\subsection*{Why the large wins concentrate on interpreter-shaped wasm}
\label{sec:eval:tt:interpreter}

The largest TtV wins are not scattered across the suite at random. They concentrate on a structurally distinct subset of the eight extension workloads---\texttt{spidermonkey-json}, \texttt{spidermonkey-markdown}, \texttt{spidermonkey-regex}, \texttt{tinygo-json}, \texttt{sqlite3}, \texttt{tract-onnx-image-classification}---and the structural reason matters for how the result generalizes. Each of these modules embeds a bytecode interpreter, a reflection-driven dispatch loop, or a graph of large tensor operators, compiled to wasm at the source toolchain's highest optimization level. SpiderMonkey runs JavaScript through its C++ bytecode dispatch loop, with the SpiderMonkey JIT tier disabled. SQLite runs SQL through its VDBE virtual machine---itself a computed-goto dispatch loop in C, pushed through Emscripten. TinyGo's \texttt{encoding/json.Unmarshal} dispatches every field decoder through \texttt{call\_indirect}, resolved at run time from \texttt{reflect.Type} tables.
The wasm-side hot path on each is opcode fetch followed by an indirect branch (or, on the tract operators, a single large function body whose tensor schedule is already fixed). The tier-2 backend cannot accelerate that hot path from already-optimized wasm bytecode: the inlining, vectorization, and dispatch choices are baked in at the source-to-wasm compile and not recoverable by re-optimizing the resulting wasm. The steady-state \texttt{LLVM/t2} ratio on these kernels is therefore flat by construction (Appendix~A.3 documents the per-kernel structural reasons and the measured ratios).

The LLVM JIT \emph{compile} cost on the same modules is not flat. It scales with module size, and these modules are large: 90\,s for each of the three SpiderMonkey kernels, 136\,s for the tract ONNX classifier, 16\,s for SQLite, 10\,s for the TinyGo JSON parser. The tiered architecture exists for this asymmetry. The user does not pay for steady-state speedup the backend cannot deliver. The architecture removes the dominant cold-start cost instead. That cost would otherwise gate any use of these modules in an edge or serverless deployment, where a cold-started JavaScript engine, SQL engine, or ONNX runtime is the actual unit of work.

The benefit is smallest on the long-running synthetic kernels, where LLVM JIT's compile cost is amortized over enough wasm execution that the wall-clock gap closes. \texttt{shootout-ctype} \texttt{TtV} is 0.93$\times$, \texttt{shootout-sieve} 0.84$\times$, \texttt{shootout-fib2} 0.90$\times$: tier-2 is slightly slower end-to-end on these kernels because the steady-state regression is no longer offset by a meaningful cold-start saving. The bimodal split between the compile-dominated and work-dominated kernels is the central observation of this evaluation. The tiered architecture buys reduced cold-start latency at a modest cost in steady-state throughput. The trade-off pays off most clearly on the real-world workloads that motivate the architecture.
