% !TeX root = ../main.tex

\begin{abstract}

WebAssembly 已廣泛部署於邊緣運算與無伺服器運算環境，這類場景對於冷啟動延遲特別敏感，因為其直接影響使用者感受到的回應速度。WasmEdge 為當前主流的伺服器端 WebAssembly 執行時期，目前僅以 LLVM 作為其唯一的 AOT 與 JIT 編譯器。儘管 LLVM 能產生高度最佳化的機器碼，其編譯流程具有顯著的延遲與記憶體開銷，並不適合資源受限的邊緣裝置或生命週期短的函式。WasmEdge 因此存在結構性的冷啟動瓶頸：每次啟動皆須先承擔完整的 LLVM 編譯成本，方能執行任何程式碼。

本論文提出一套適用於 WasmEdge 的分層編譯架構，目的在於將啟動延遲與最終程式碼品質解耦。第一層為 baseline JIT 編譯層 (Tier-1)，於模組載入時以低成本將所有函式編譯為原生程式碼，立即提供執行能力。第二層為最佳化編譯層 (Tier-2)，重用 WasmEdge 既有的 LLVM 前端，透過合成 mini-module、三類 ABI thunk 橋接機制，以及納入靜態熱點呼叫對象的批次組合策略，選擇性地重新編譯熱點函式。第三項為在堆疊替換 (On-Stack Replacement, OSR) 機制，於最外層迴圈反向邊插入監測點，將正在執行中的 tier-1 框架於迴圈執行過程中遷移至 tier-2，解決函式入口替換在「單次呼叫者」場景下無法加速的問題。

本架構於 WasmEdge 中實作，並以強化後的 sightglass 測試套件進行評估。在 39 個分析測試項目上，本架構將編譯時間相較於完整模組 LLVM-JIT 縮短了幾何平均 11.1 倍，同時於穩態吞吐量上仍保留了約 86\% 的 LLVM-JIT 表現。端到端執行時間因此平均縮減 1.73 倍，於最大型工作負載上更展現出數量級的改進：\texttt{tinygo-json} 達 21.9 倍、ONNX 分類器達 14.1 倍、三個 SpiderMonkey 測試項目皆達約 10 倍。OSR 機制亦於單次呼叫者場景中彌補殘餘差距---將 \texttt{shootout-ctype} 由 0.37 倍提升至 0.88 倍、\texttt{blind-sig} 由 0.40 倍提升至 0.80 倍（相對於 LLVM-JIT）。

\end{abstract}

\begin{abstract*}

WebAssembly is increasingly deployed in edge computing and serverless environments, where cold-start latency directly determines user-perceived responsiveness. WasmEdge, a leading server-side WebAssembly runtime, relies solely on LLVM for both ahead-of-time (AOT) and just-in-time (JIT) compilation. Although LLVM produces highly optimized machine code, its compilation pipeline incurs compile-time and memory overheads that are unsuitable for resource-constrained edge devices and short-lived functions. Every invocation pays the full LLVM compile cost before any code executes.

This thesis presents a tiered compilation architecture for WasmEdge that decouples cold-start latency from peak code quality. A baseline JIT tier compiles every function eagerly at module load time. An optimizing tier reuses the existing WasmEdge LLVM frontend to selectively recompile hot functions through mini-module synthesis, three ABI bridge thunks, and a batch composition policy that admits statically-hot direct callees. An on-stack replacement (OSR) mechanism instruments outermost-loop back-edges in baseline-tier code and migrates currently-running frames into the optimizing tier mid-execution, addressing the one-shot-caller scenario in which function-entry promotion alone delivers no speedup.

The architecture is implemented in WasmEdge and evaluated on a strengthened variant of the sightglass benchmark suite. Across 39 analyzed kernels, the architecture cuts compile time by a geometric-mean factor of 11.1$\times$ over whole-module LLVM JIT while preserving 86\% of LLVM-JIT steady-state throughput. End-to-end wall-clock time improves by 1.73$\times$ on average, with order-of-magnitude wins on the largest workloads: 21.9$\times$ on \texttt{tinygo-json}, 14.1$\times$ on the ONNX classifier, and approximately 10$\times$ on each of the three SpiderMonkey kernels. On-stack replacement closes the residual gap on one-shot-caller workloads, raising \texttt{shootout-ctype} from 0.37$\times$ to 0.88$\times$ and \texttt{blind-sig} from 0.40$\times$ to 0.80$\times$ relative to LLVM JIT.

\end{abstract*}
